---
title: "Optimal proposed duration of a TB prophylactic treatment"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = TRUE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      message    = FALSE,
                      warning    = FALSE,
                      dev.args   = list(pointsize = 11),
                      fig.height = 3.5,
                      fig.width  = 4.24725,
                      fig.retina = 2,
                      fig.align  = "center")
```

## 1 Overview

### 1.1 Context and question
The goal is to reduce TB prevalence in the population by proposing a
prophylactic treatment of a given duration to people recently exposed to a
confirmed TB case. The shorter the proposed duration, the lower the efficacy
of the treatment but, also -- we hypothesize --, the higher the uptake rate. The 
question then is whether a higher uptake rate can epidemiologically
(over-)compensate at the population level a lower efficacy of the treatment at
the individual level. If yes, then we are also interested in knowing what is the
proposed duration of the treatment that would lead to the lowest TB prevalence
in the population.

### 1.2 Method

The model contains 2 sub-models. The first one is an epidemiological model of TB
transmission in the population from which we derive a formula of the
prevalence of TB in the population. The second one is a model of prophylactic
treatment policy that accounts for

* contact tracing efficacy;
* treatment uptake as a function of proposed treatment duration;
* the effective treatment duration as a function of the proposed treatment
  duration;
* treatment efficacy as a function of the effective treatment duration.

The distribution of the effective treatment duration as a function of the
proposed treatment duration is itself derived from assumptions made on how
treatment adherence depends on the proposed treatment duration.

The efficacy of the prophylactic treatment policy is then integrated into the
formula of the TB prevalence in order to derive the relationship between the
proposed duration of the prophylactic treatment and the TB prevalance in the
population.

#### 1.2.1 Epidemiological sub-model
We develop a simple 5-parameters epidemiological model of TB that considers 3
clinical status (**Figure 1**): 

* susceptible ($S$);
* infected but neither sick nor infectious ($I$);
* sick and infectious ($D$).

In this model people can transit between $S$ and $I$ and between $I$ and $D$ in
both directions, and from $D$ to $I$. Formulas of the prevalences of $I$ and $D$
at the epidemiological equilibrium are derived. The model is qualibrated by using 
information on

* the prevalence of $I$;
* the prevalence of $D$;
* the proportion of infected that will ultimately develop the disease;
* the proportion of people developing the disease that do so within a given time;
* the proportion of relapse.

#### 1.2.2 Prophylactic treatment sub-model

Here the sub-model aims at computing the efficacy of the prophylactic treatment
policy by accounting for

* contact tracing efficacy;
* treatment uptake;
* treatment adherence that determines the effective treatment duration upon
  uptake;
* treatment efficacy that depends on the effective treatment duration.

**Treatment uptake:** in absence of detailed information of people's behaviour
related to treatment, we phenomenologically model the probability of treatment
uptake as a function of the proposed duration of the proposed prophylactic
treatment. For that, we use a 3-parameter Hill equation that provides a S-curve
with great shape flexibility (**Figure 4**).

**Treatment adherence and effective treatment duration:** any person taking the
treatment can drop it at any time. We thus model the treatment adherence as a
function of the treatment duration in order to compute the effective duration of
treatment that, accounting for drop-outs, will be lower than the proposed
treatment duration. To do so, we consider that any given day in the treatment
the probability of dropping it increases with the number of days since the start
of the treatment (reflecting a *"fatigue"* of taking the treatment) and, at the
same time, decreases with the number of days left in the treatment (reflecting
the fact that *somebody is not likely to give up that close to the "finish line"*). 
(An alternative to this second probability could be to express the time left in
the treatment as a proportion of the total duration of the treatment.) Each of
these 2 effects is modelled by a 3-parameter Hill equation as described above
and, from these two probabilities combined, we derive the probability
distribution of the effective treatment duration among the people taking the
treatment (**Figure 5**). This probability distribution depends on the proposed 
duration of the treatment. From here we can either computer the mean of the
distribution for the rest of the analysis or, instead, propagate this
distribution through the rest of the analysis.

**Treatment efficacy:** the last piece of the puzzle consists in modelling the
treatment efficacy as a function of its effective duration and here again we
make use of a 3-parameter Hill equation to do so (**Figure 6**).

#### 1.2.3 TB prevalence as a function of proposed treatment duration

From the above section we have the efficacy of the prophylactic treatment policy
as a function of its proposed duration. Next step consists in integrating this
efficacy into the formula of the TB prevalence by converting this efficacy into
a rate accounting for the competing risks as defined by the epidemiological
model.Once this is done, we can express the prevalence of TB directly as a
function of the proposed duration of the proposed prophylactic treatment and
then look whether there is a value of the duration of the duration of the
prophylactic treatment that minimizes the TB prevalence (**Figure 7**). As
mentioned above, this can be done by considering either the mean effective
duration of treatment, or the whole distribution of the effective duration of
treatment.

## 2 Packages

The package used in this analysis:

```{r}
library(tibble)
library(deSolve)
library(purrr)
library(dplyr)
library(parallel)
```

## 3 Functions

Tuning and defining some utilitary functions.

```{r}
lwd_val <- 2

seq2 <- function(...) seq(..., le = 512)

plot2 <- function(..., col = 4) plot(..., type = "l", lwd = lwd_val, col = col)

plot3 <- function(...) plot2(..., xlab = "time (years)")

plot4 <- function(...) plot2(..., xlab = "actual duration of treatment (days)")

lines2 <- function(..., col = 2) lines(..., lwd = lwd_val, col = col)

legend2 <- function(...) legend(..., lty = 1, lwd = lwd_val, bty = "n")

abline2 <- function(..., col = 2) abline(..., lwd = lwd_val, col = col)

mclapply2 <- function(...)
  parallel::mclapply(..., mc.cores = parallel::detectCores() - 1) 
```

A few utilitary functions:

```{r}
polygon2 <- function(x, y1, y2, col = 4, alpha = .2, ...) {
  polygon(c(x, rev(x)), c(y1, rev(y2)), border = NA,
          col = adjustcolor(col, alpha), ...)
}
```

```{r}
polygon3 <- function(x, y, col = 4, alpha = .2, ...) {
  polygon(c(x[1], x, tail(x, 1)), c(0, y, 0), col = adjustcolor(col, alpha),
          border = NA)
}
```

A function that will draw from the probability distribution:

```{r}
draw <- function(probs, n) {
  rep(seq_along(probs), as.vector(rmultinom(1, size = n, prob = probs)))
}
```

```{r}
get_val <- function(ind, vect) vect[ind]
```


## 4 Epidemiological model

### 4.1 Assumptions

* constant population size
* frequency-dependence transmission
* no mortality
* no immunity against TB

### 4.2 Epidemiological framework

<div style="text-align:center"><img src ="SID.png" width="160"/></div>

**Figure 1:** flow chart of the epidemiological model in which people can be in
3 clinical status: non-infected and susceptible ($S$), infected but neither
sick nor infectious ($I$) and diseased people ($D$).

#### 4.2.1 Parameters


* $N$: population size (ind)
* $\beta$: per capita infectious contact rate (/year/ind)
* $\gamma$: per capita rate of clearance of non-diseased, either naturally or
from whatever treatment policy currently in place (/year)
* $\delta$: per capita rate of clearance of diseased, either naturally or
from whatever treatment policy currently in place (/year)
* $\sigma$: per capita rate of developing disease once infected (/year)
* $q$: proportion of treated diseased that do not clear the bacillus
* $\pi(d)$: per capita rate of clearance of non-diseased, due to prophylactic
treatment (/year). Depends the on duration $d$ of prophylactic treatment (see
below)


#### 4.2.2 Epidemiological dynamics:

$$
\frac{dS}{dt} = (1-q)\delta D + (\gamma + \pi(d)) I - \beta\frac{D}{N}S
$$

$$
\frac{dI}{dt} = \beta\frac{D}{N}S + q\delta D - (\sigma + \gamma + \pi(d)) I
$$

$$
\frac{dD}{dt} = \sigma I - \delta D
$$

#### 4.2.3 Constant population size:

$$
N = S + I + D
$$

#### 4.2.4 Equilibrium 

$$
D^* = \frac{\beta - \left(1 - q + \frac{\gamma + \pi}{\sigma}\right)\delta}
{\left(1 + \frac{\delta}{\sigma}\right)\beta}N
$$


$$
I^* = \frac{\delta}{\sigma}D^*
$$

$$
S^* = N - I^* - D^*
$$


#### 4.2.5 Numerical verification

The model in R:

```{r}
model <- function(S0, I0, D0, beta, sigma, gamma, delta, q, pi, times) {
  N <- S0 + I0 + D0
  c(S = S0, I = I0, D = D0) |> 
    ode(times,
        function(time, state, pars) {
          with(as.list(c(state, pars)), {
            infections <- beta * D * S / N
            dS <- (1 - q) * delta * D + (gamma + pi) * I - infections
            dI <- infections + q * delta * D - (sigma + gamma + pi) * I
            dD <- sigma * I - delta * D
            list(c(dS, dI, dD))
          })},
        c(beta = beta, sigma = sigma, gamma = gamma, delta = delta, q = q, pi = pi)) |>
    as.data.frame() |> 
    as_tibble()
}
```

The equilibrium values in R:

```{r}
d_star <- function(gamma, sigma, delta, beta, q, pi, N) {
  N * (beta - delta * (1 - q + (gamma + pi) / sigma)) /
    (beta * (1 + delta / sigma))
}

i_star <- function(gamma, sigma, delta, beta, q, pi, N) {
  delta * d_star(gamma, sigma, delta, beta, q, pi, N) / sigma
}

s_star <- function(gamma, sigma, delta, beta, q, pi, N) {
  N - i_star(gamma, sigma, delta, beta, q, pi, N) -
    d_star(gamma, sigma, delta, beta, q, pi, N)
}
```

Let's compare:

```{r fig.width = 7, fig.height = 5 / 2, margin = FALSE}
S0 <- 1e6 - 10 # ind
I0 <- 0 # ind
D0 <- 10 # ind
beta <- 1 # /year/ind
sigma <- 1.15 # /year
gamma <- .1 # /year
delta <- .1 # /year
q <- .5
pi <- 0 # in absence of prophylaxis

sims <- model(S0, I0, D0, beta, sigma, gamma, delta, q, pi, seq2(0, 50))

N <- S0 + I0 + D0

opar <- par(mfrow = c(1, 3))
with(sims, plot3(time, S, ylab = "susceptibles S"))
abline2(h = s_star(gamma, sigma, delta, beta, q, pi, N))
with(sims, plot3(time, I, ylab = "infected non-diseased I"))
abline2(h = i_star(gamma, sigma, delta, beta, q, pi, N))
with(sims, plot3(time, D, ylab = "diseased D"))
abline2(h = d_star(gamma, sigma, delta, beta, q, pi, N))
par(opar)
```

**Figure 2:** numerical verification of the formulas for the equilibrium values
of $S$, $I$ and $D$: numerical simulation in blue, values from the formulas in
red.

#### 4.2.6 Estimating / assessing parameters

If $p\%$ of those $I$ who will become $D$ within $E$ years, then the $\sigma$
rate should read

$$
\sigma = -\frac{\log(1 - p)}{E}
$$

If $x$ (typically $3$ to $5\%$) is the proportion of $I$ that will ultimately
move to $D$, then, we have

$$
\frac{\sigma}{\sigma + \gamma} = x
$$

which leads to this expression for $\gamma$:

$$
\gamma = \frac{1 - x}{x}\sigma
$$

We are left with two parameters to estimate ($\beta$ and $\delta$) that we can
estimate from observed values of $I^*$ (between 10 and 40%) and $D^*$ (between
0.15 and 0.30%). First $\delta$:

$$
\delta = \sigma\frac{I^*}{D^*}
$$

And $\beta$:

$$
\beta = \frac{(1 - q)\delta D^* + \gamma I^*}{(N - I^* - D^*)D^*}N
$$
In R:

```{r}
parameters_values <- function(
    p = .9, E = 2, # p% of people developing disease doing so within E years
    x = .04, # proportion of I that will become D (3 to 5%)
    I = .25, # prevalence of I (between 10 and 40%)
    D = 450 / 100000, # prevalence of D (between 150 and 300 / 100,000)
    q = .15) { # proportion of "recovered" D that actually go to I instead of S
  
  sigma <- - log(1 - p) / E
  gamma <- sigma * (1 - x) / x
  delta <- sigma * I / D
  beta <- ((1 - q) * delta * D + gamma * I) / ((1 - I - D) * D)
  
  c(beta = beta, gamma = gamma, sigma = sigma, delta = delta)
}
```

```{r}
parameters_values()
```

```{r fig.width = 7, fig.height = 5 / 2, margin = FALSE}
q <- .15
D0 <- 10 # ind
S0 <- 1e6 - D0 # ind
I0 <- 0 # ind

N <- S0 + I0 + D0
parms <- parameters_values(q = q)
beta  <- parms["beta"] # /year/ind
sigma <- parms["sigma"] # /year
gamma <- parms["gamma"] # /year
delta <- parms["delta"] # /year

sims <- model(S0, I0, D0, beta, sigma, gamma, delta, q, 0, seq2(0, 5))

opar <- par(mfrow = c(1, 3))
with(sims, plot3(time, S, ylab = "susceptibles S"))
abline2(h = s_star(gamma, sigma, delta, beta, q, pi, N))
with(sims, plot3(time, I, ylab = "infected non-diseased I"))
abline2(h = i_star(gamma, sigma, delta, beta, q, pi, N))
with(sims, plot3(time, D, ylab = "diseased D"))
abline2(h = d_star(gamma, sigma, delta, beta, q, pi, N))
par(opar)
```

**Figure 3:** same as Figure 2 but with realistic values of the parameters,
verifying that the prevalences of $I$ and $D$ fit the expected values.

## 5 Prophylaxis model

### 5.1 The Hill equation

In what follows we will model many phenomena phenomenologically using the
3-parameter monotonically increasing Hill equation, the general equation of
which is

$$
y = Y\frac{x^h}{X^h + x^h}
$$
where $Y$ is the maximum value that $y$ can take, $X$ is the value of $x$ at
which $y$ reaches half of its maximum value $Y$ and $h$ is the Hill coefficient
that controls the shape of the relationship with S-shape for $0 < h < 1$ and a
simple saturating shape when $h \ge 1$.

```{r}
hill <- function(x, X, Y, h) {
  x2h <- x^h
  Y * x2h / (X^h + x2h)
}
```

Let's illustrate the properties of this equation:

```{r}
xs <- seq2(0, 40)
X <- 15
Y <- .8

plot(NA, xlim = c(0, max(xs)), ylim = c(0, Y),
     xlab = "x values", ylab = "y values")
hs <- exp(seq(-3, 3))
abline2(v = X, col = "grey")
abline2(h = Y, col = "grey")
walk2(hs, rev(RColorBrewer::brewer.pal(n = length(hs), "Spectral")),
      ~ lines2(xs, hill(xs, X, Y, h = .x), col = .y))
```


### 5.2 Treatment uptake

We can model the probability of treatment uptake as a function of the treatment
duration using a Hill equation as so:

```{r}
ds <- seq2(0, 30)
plot2(ds, 1 - hill(ds, 15, 1, 4), ylim = 0:1,
      xlab = "duration of treatment (days)", ylab = "uptake probability")
```

**Figure 4:** modelled probability of treatment uptake as a function of proposed
treatment duration.


### 5.3 Treatment adherence and effective duration of treatment

Let's assume that the probability $f$ that somebody stops his/her treatment
increases with the number of days s/he's been taking the treatment according to
a Hill equation, reflecting some fatigue effect of being in the treatment. Let's
further assume that this probability is gets mitigated the more we get close to
the end of the treatment, reflecting the effect that a person is not likely to
drop a treatment anymore when s/he gets close to the end of the treatment. This
mitigation effect $m \in [0,1]$ is modelled by with a Hill equation as well. The
probability $p(t)$ that somebody on his/her day $t$ of treatment drops it then
reads

$$
p(t) = m(t) \times f(t)
$$

From this we can express the probability that somebody stops his/her treatment
**at time $t$** as

$$
P(t) = p(t)\prod_{x=0}^{{}^{-}t}(1 - p(x))^{dx}
$$

The probability distribution of the actual durations of treatment (*aka*
**effective durations of treatment**) can then be expressed by:

$$
\varphi(t, d) = \frac{P(t)}{\int_{0}^d P(x)dx}
$$

where $d$ is the **proposed duration of treatment**.

Let's look at some examples:

```{r}
probabilities <- function(
    pd = 40, # propsed duration of treatment
    Xf = 25, Yf = .8, hf = 7, # fatigue
    Xmp = .875, Ym = 1, hm = 50, # mitigation of fatigue
    lgnd = TRUE) { # whether to add the legend to the plot or not
  
  Xm <- Xmp * pd
  ts <- seq2(0, pd)

  fatigue <- hill(ts, Xf, Yf, hf)
  mitigation <- 1 - hill(ts, Xm, Ym, hm)
  
  plot2(ts, fatigue, ylim = 0:1, col = 3,
        xlab = "time in treatment (days)",
        ylab = "probability or mitigation factor")
  lines2(ts, mitigation)
  lines2(ts, mitigation * fatigue, col = 4)
  
  if (lgnd)
    legend2("topleft", col = c(3, 2, 4),
            legend = c("fatigue", "mitigation of fatigue", "dropping probability"))
}
```

Executing the function:

```{r}
probabilities()
```

Let's now work out the probability distribution of the effective durations of
treatment:

```{r}
dropping_probability <- function(
    pd = 40, # proposed duration of treatment
    Xf = 25, Yf = .8, hf = 7, # fatigue
    Xmp = .875, Ym = 1, hm = 50, # mitigation of fatigue
    by = pd / (le - 1), le = 512) { # sampling day in the treatment

  ts <- seq(0, pd, by)
  tibble(x = ts,
         y = hill(ts, Xf, Yf, hf) * (1 - hill(ts, Xmp * pd, Ym, hm)))
}
```



```{r}
effective_duration <- function(
    pd = 40, # proposed duration of treatment
    Xf = 25, Yf = .8, hf = 7, # fatigue
    Xmp = .875, Ym = 1, hm = 50, # mitigation of fatigue
    by = pd / (le - 1), le = 512) { # sampling effective durations of treatment
  
  dp <- dropping_probability(pd, Xf, Yf, hf, Xmp, Ym, hm, by, le)
  y <- dp$y
  dens <- c(y[1], y[-1] * cumprod(1 - y[-length(y)]))
  tibble(x = dp$x,
         y = dens / sum(dens))
}
```

Let's try it:

```{r}
with(effective_duration(), {
  plot4(x, y, ylab = "probability density")
  polygon3(x, y)
})
```



```{r}
pd <- 40
Xf <- 100
Yf <- .8
hf <- 7
Xmp <- .875
Ym <- 1
hm <- 50
by <- .1
proposed_durations <- c(20, 40, 60, 70, 80)
color_values <- seq_along(proposed_durations) + 1

eds <- map(proposed_durations,
           ~ effective_duration(.x, Xf, Yf, hf, Xmp, Ym, hm, by))

plot(NA,
     xlim = c(0, max(proposed_durations)),
     ylim = c(0, max(unlist(map(eds, ~ .x$y)))),
     xlab = "effective duration of treatment (days)",
     ylab = "probability density")

walk2(eds, color_values, ~ with(.x, {
  lines2(.x$x, .x$y, col = .y)
  polygon3(x, y, col = .y)
}))

abline2(v = proposed_durations, col = color_values)
```


### 5.4 Treatment efficacy

We can assume that the treatment efficiency as a function of the actual duration
of treatment also follows a Hill equation:

```{r}
ts <- seq2(0, 30)
plot4(ts, hill(ts, 10, 1, 1), ylim = 0:1, ylab = "treatment efficacy")
```

**Figure 6:** modelled treatment efficacy as a function of the effective
treatment duration.

### 5.5 Putting everything together

By putting everything together, we can express the probability of successful
prophylactic treatment of infected as

$$
\chi(d) = \tau \times \upsilon(d) \times \varepsilon(\delta(d)) 
$$

where

* $\tau$ is the proportion of infected that can be identified through contact
tracing;
* $\upsilon(d)$ is the treatment uptake as a function of the proposed
treatment duration $d$;
* $\delta(d)$ is the effective treatment duration as a function of the proposed
treatment duration $d$;
* $\varepsilon(\delta(d))$ is the treatment efficacy as a function of the
effective treatment duration $\delta(d)$.

From this, it comes that the rate $\pi(d)$ can be expressed as

$$
\pi(d) = \frac{\chi(d)}{1 - \chi(d)}(\sigma + \gamma)
$$

This expression of $\pi(d)$ can be fed into the expression of $D*$ to get the
effect of the duration of the proposed prophylactic treatment on the number $D*$
of people with TB:

$$
D^*(d) = \frac{\beta - (1 - q + (\gamma + \pi(d)) / \sigma)\delta}
{(1 + \delta / \sigma)\beta}N
$$


```{r}
duration2incidence <- function(
    ed, # effective duration
    pd = 40, # proposed treatment duration
    tau = .45, # contact tracing efficacy (30-60%)
    Xu = 15, Yu = 1, hu = 4, # treatment uptake
    Xe = 10, Ye = 1, he = 1, # treatment efficacy
    gamma, sigma, delta, beta, # epidemiological parameters
    q = .15, # proportion of "recovered" D that actually goes to I instead of S
    N = 1e5) {  # population size
  
  chi <- tau * (1 - hill(pd, Xu, Yu, hu)) * hill(ed, Xe, Ye, he)
  d_star(gamma, sigma, delta, beta, q, chi * (sigma + gamma) / (1 - chi), N)
}
```


```{r}
tb_cases <- function(
    pd = 40, # proposed treatment duration
    tau = .45, # contact tracing efficacy (30 to 60%)
    Xu = 15, Yu = 1, hu = 4, # treatment uptake
    Xf = 25, Yf = .8, hf = 7, # fatigue effect
    Xmp = .875, Ym = 1, hm = 50, # mitigation effect
    Xe = 10, Ye = 1, he = 1, # treatment efficacy
    p = .9, E = 2, # p% of people developing disease doing so within E years
    x = .04, # proportion of I that will become D (3 to 5%)
    I = .25, # prevalence of I (between 10 and 40%)
    D = 450 / 200000, # prevalence of D (150 to 300 / 100,000)
    q = .15, # proportion of "recovered" D that actually goes to I instead of S
    N = 1e5, # population size
    n = 1e5, # number of draws to sample the distribution of the effective durations
    by = pd / (le - 1), le = 1e4) { # integration
  
  parms <- parameters_values(p, E, x, I, D, q)
  sigma <- parms["sigma"]
  gamma <- parms["gamma"]
  
  ts <- seq(0, pd, by)
  drop_prob <- hill(ts, Xf, Yf, hf) * (1 - hill(ts, Xmp * pd, Ym, hm))

  draws <- c(drop_prob[1],
             drop_prob[-1] * cumprod(1 - drop_prob[-length(drop_prob)])) |>
    draw(n) |> 
    table()
  
  draws |> 
    names() |> 
    as.numeric() |> 
    get_val(ts) |> 
    map_dbl(~ duration2incidence(.x, pd, tau, Xu, Yu, hu, Xe, Ye, he, gamma,
                                 sigma, parms["delta"], parms["beta"], q, N)) |> 
    rep(draws)
}
```



```{r}
xs <- seq2(1, 100)
probs <- 0:10 / 10

ys <- xs |>
  mclapply2(function(x) quantile(tb_cases(x), probs)) |> 
  bind_rows()
```

```{r}
plot(NA, xlim = c(0, max(xs)), ylim = c(0, max(ys)))
nbcol <- length(probs)
i <- (nbcol - 1) /2
walk(1:i, ~ polygon2(xs, ys[[.x]], ys[[nbcol - .x]]))
lines(xs, ys[[i + 1]], col = 4)
```

**Figure 7:** tuberculosis prevalence as a function of proposed duration of
prophylactic treatment.
