---
title: "Model v2.0"
output:
  html_document:
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = TRUE,
                      cache      = TRUE,
                      autodep    = TRUE,
                      message    = FALSE,
                      warning    = FALSE,
                      dev.args   = list(pointsize = 11),
                      fig.height = 3.5,
                      fig.width  = 4.24725,
                      fig.retina = 2,
                      fig.align  = "center")
```

**Variables:**

* $S$: non-infected
* $I_1$: recently infected
* $I_2$: infected a long time ago
* $I_3$: treated but not sterilized
* $D$: sick and infectious
* $N = S + I_1 + I_2 + I_3 + D$

**Parameters:**

* $\beta$: infectious contact rate (/ind/year)
* $\sigma_1$ and $\sigma_2$: clearance rates (/year) of $I_1$ and $I_2$
respectively, either through natural immunity or prophylactic treatment
* $\delta_1$, $\delta_2$ and $\delta_3$: progression rates (/year) of infection
to disease for $I_1$, $I_2$ and $I_3$ (relapse) respectively
* $\gamma$: rate (/year) of recovery with clearance either from treatment or
natural immunity
* $\rho$: rate (/year) of recovery without clearance (ultimately leading to
relapse)

**Note:** all these rates are *per capita*

**Flow diagram:**

<div style="text-align:center"><img src ="SI1I2I3D.png" width="280"/></div>

**Differential equations:**

$$
\begin{align} 
  \frac{dS}{dt}   &= \sigma_1 I_1 + \sigma_2 I_2 + \gamma D - \beta \frac{D}{N} S \\
  \frac{dI_1}{dt} &= \beta \frac{D}{N} S - (\sigma_1 + \delta_1 + \tau) I_1       \\
  \frac{dI_2}{dt} &= \tau I_1 - (\delta_2 + \sigma_2) I_2                         \\
  \frac{dI_3}{dt} &= \rho D - \delta_3 I_3                                        \\
  \frac{dD}{dt}   &= \delta_1 I_1 + \delta_2 I_2 + \delta_3 I_3 - (\gamma + \rho) D
\end{align}
$$
**Endemic equilibrium:**

$$
\begin{align}
  A     &= \frac{(\sigma_2 + \delta_2) \gamma}
                {(\sigma_2 + \delta_2) \delta_1 + \tau \delta_2}     \\
  B     &= 1 + \frac{(\sigma_2 + \delta_2 + \tau) \gamma}
                    {(\sigma_2 + \delta_2) \delta_1 + \tau \delta_2}
             - \frac{\rho}{\delta_3}                                 \\
  D^*   &= \left(1 - \frac{\sigma_1 + \delta_1 + \tau}{\beta}\right)
           \frac{A}{B} N                                             \\
  I_1^* &= AD^*                                                      \\
  I_2^* &= \frac{\tau}{\sigma_2 + \delta_2} I_1^*                    \\
  I_3^* &= \frac{\rho}{\delta_3}D^*                                  \\
  S^*   &= N - I_1^* - I_2^* - I_3^* - D^*
\end{align}
$$

**R code for the differential equations:**

```{r}

```

