---
title: "Of mice and men"
output:
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      message = FALSE,
                      cache   = TRUE)
```


Looking at data from
Zhang *et al.*
Short-course chemotherapy with TMC207 and rifapentine in a murine model of
latent tuberculosis infection.
*American Journal of Respiratory and Critical Care Medicine* 184.6 (2011): 732-737.
[10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)

The data show the proportion of positive mice as a function of time. What we are
interested in is the CFU clearance dynamics at the individual level. Since the
former is a consequence of the latter, we develop a simple model of CFU
clearance at the individual level from which we compute the consequences at
(small-size) population level. By confronting the model-predicted consequences
at population level with actual experimental data, we are able to estimate the
two parameters of the model. Once the model calibration is done, we can use the
model to explore what would be the effect of decreasing the initial value of CFU
on the time at which the individual clear the infection.

The model itself is a simple exponential decay model with inter-individual
variability on the rate of decay that is modelled by a Gamma distribution.

## 1 Packages

Loading the packages needed for the analysis:

```{r}
library(readxl)
library(dplyr)
library(purrr)
library(tidyr)
library(bbmle)
```


## 2 Utilitary functions

A function that adds proportions estimates and confidence intervals to a data
frame from a column of successes and a column or trials:

```{r}
add_prop_est <- function(df, x, n, p = "est", l = "lwr", u = "upr", ...) {
  df |> 
    mutate(test   = map2({{ x }}, {{ n }}, prop.test, ...),
           "{p}" := map_dbl(test, ~ .x[["estimate"]]),
           conf   = map(test, ~ setNames(.x[["conf.int"]], c(l, u)))) |> 
    unnest_wider(conf) |> 
    select(- test)
}
```

A reparameterization of `dgamma()`:

```{r}
dgamma2 <- function(x, mu, sigma) {
  dgamma(x, mu^2 / sigma^2, scale = sigma^2 / mu)
}
```

A reparameterization of `qgamma()`:

```{r}
qgamma2 <- function(p, mu, sigma) {
  qgamma(p, mu^2 / sigma^2, scale = sigma^2 / mu)
}
```


## 3 The data

Reading the data:

```{r}
table2 <- read_excel("table 2 Zhang et al 2011.xlsx")
```

The data look like this:

```{r}
table2
```

Preparing the data for vizualization:

* jittering the duration variable slightly in order to avoid visual overlap)
* computing the proportions of positive mice with 95% confidence intervals

```{r}
table2a <- table2 |>
  add_prop_est(positive, total) |> 
  mutate_at("duration", jitter, factor = .2)
```

It looks like this:

```{r}
table2a
```

A function that adds proportion estimates and confidence intervals to a plot:

```{r add_data}
add_data <- function(x, col) {
  table2a |> 
    filter(group == x) |> 
    with({
      points(duration, est, col = col, type = "o")
      arrows(duration, lwr, duration, upr, .1, 90, 3, col)
      lines(duration, lwr, lty = 2, col = col)
      lines(duration, upr, lty = 2, col = col)
    })
}
```

The plot showing points estimates with confidence intervals from the data:

```{r dependson = "add_data"}
treatments <- unique(table2$group)

plot(NA, xlim = c(0, 6), ylim = 0:1,
     xlab = "duration of treatment (months)",
     ylab = "proportion of positive mice")

walk2(treatments, 1:4, add_data)

legend("bottomright", legend = treatments, col = 1:4, lwd = 1, bty = "n")
```

## 4 A simple model of decay

### 4.1 Within-individual exponential decay of CFU

Let's assume that the CFU in an individual decays at a constant rate $\lambda$.
Then CFU as a function of time $t$ reads:

$$
\mbox{CFU}(t) = \mbox{CFU}_0 e^{-\lambda t}
$$

A function that gives CFU as a function of initial CFU$_0$, decay rate and time:

```{r}
cfu <- function(cfu0, lambda, t) {
  cfu0 * exp(- lambda * t)
}
```

Let's try it:

```{r}
cfu(1e4, .115, 30)
cfu(1e4, .115, c(30, 60))
```

### 4.2 Inter-individual variability of decay rate

The next step is to consider that the rate of exponential decay varies slightly
from individual to individual. Let's consider that this inter-individual
variability follows a Gamma distribution. Next, we want to work out how this
inter-individual variability translates into a proportion of individuals with at
least 1 CFU in a population of an infinite number of individuals at time $t$.
This is done looking at the cumulative probability of the Gamma distribution of
the decay rate that leads to the quantile of the inter-individual CFU
distribution that is the closest to 1. To do so, we need a function that
expresses the distance of the quantile of the inter-individual CFU distribution
to 1:

```{r}
distance_to_1 <- function(p, lambda, sigma, N0, t) {
  abs(1 - cfu(N0, qgamma2(p, lambda, sigma), t))
}
```

We then use this function in the following function that converts values for
$\lambda$ and $\sigma$ of the Gamma distribution of the inter-individual
variability of exponential decay rate into the probability parameter of a
binomial distribution that corresponds to the expected proportion of individuals
in the population with CFU > 1:

```{r}
probability <- function(lambda, sigma, N0, t, epsilon = 1e-16) {
  upr <- cfu(N0, qgamma2(epsilon, lambda, sigma), t)
  if (upr < 1) {
    return(0)
  } else {
    lwr <- cfu(N0, qgamma2(1 - epsilon, lambda, sigma), t)
    if (lwr > 1) {
      return(1)
    }
    else {
      optimize(distance_to_1, 0:1, lambda = lambda, sigma = sigma, N0 = N0, t = t)$minimum
    }
  }
}
```

Let's try it:

```{r}
xs <- seq(.1, .5, le = 512)
plot(xs, map_dbl(xs, probability, sigma = .011, N0 = 1e4, t = 30, epsilon = 1e-16),
     type = "l", col = 4, lwd = 3, xlab = "sigma", ylab = "probability")
```


## 5 Maximum-likelihood estimation of the model's parameters

The first step consists in expressing a function that computes minus log
likelihood:

```{r}
threshold <- - log(1e-16)

mLL <- function(lambda, sigma, N0, data, epsilon = 1e-16) {
  min(threshold,
      - sum(dbinom(data$positive,
                   data$total,
                   map_dbl(data$duration, ~ probability(lambda, sigma, N0, .x * 30, epsilon)),
                   TRUE)))
}
```

Let's try it:

```{r}
mLL(lambda = .3, sigma = .011, N0 = 1e4, data = filter(table2, group == "RIF"), epsilon = 1e-16)
mLL(lambda = .1, sigma = .011, N0 = 1e4, data = filter(table2, group == "RIF"), epsilon = 1e-16)
```

### 5.1 Vizualization of the likelihood surface:

A function that vizualizes the likelihood surface:

```{r}
show_mLL <- function(drug, lambda_val, sigma_val, N0 = 1e4, epsilon = 1e-16, ...) {
  values <- expand.grid(lambda_val, sigma_val)
  
  mLL_val <- map2_dbl(values$Var1, values$Var2, mLL, N0 = N0,
                      data = filter(table2, group == drug), epsilon = epsilon) |> 
    matrix(length(lambda_val))
  
  image(lambda_val, sigma_val, mLL_val, xlab = "lambda", ylab = "sigma")
  contour(lambda_val, sigma_val, mLL_val, add = TRUE, ...)
}
```

Let's try it:

```{r}
show_mLL("INH", seq(.001, .05, le = 100), seq(.0001, .012, le = 100), levels = c(.002, .01, .1))
show_mLL("RIF", seq(.001, .2, le = 100), seq(.0001, .1, le = 100), levels = 1:5)
show_mLL("RIF+INH", seq(.05, .15, le = 100), seq(.0001, .15, le = 100), levels = 1:5)
show_mLL("RPT+INH", seq(.001, .5, le = 100), seq(.0001, .1, le = 100), levels = 1:5)
```

### 5.2 MLE using the BFGS quasi-Newton algorithm

A function that performs estimations:

```{r}
estimations <- function(lambda, sigma, lower, upper, drug, N0 = 1e4, epsilon = 1e-16) {
  data <- filter(table2, group == drug)
  mle2(function(lambda, sigma) mLL(lambda, sigma, N0, data, epsilon),
       list(lambda = lambda, sigma = sigma), "L-BFGS-B",
       lower = lower, upper = upper)
}
```

Let's try it:

```{r}
est_RIF <- estimations(lambda = .08, sigma = .02,
                       lower = c(lambda = .06, sigma = .01),
                       upper = c(lambda = .09, sigma = .05),
                       drug = "RIF")
```

```{r}
est_RIFINH <- estimations(lambda = .11, sigma = .04,
                          lower = c(lambda = .09, sigma = .02),
                          upper = c(lambda = .11, sigma = .05),
                          drug = "RIF+INH")
```

```{r}
est_RPTINH <- estimations(lambda = .3, sigma = .04,
                          lower = c(lambda = .28, sigma = .01),
                          upper = c(lambda = .32, sigma = .08),
                          drug = "RPT+INH")
```

Which gives:

```{r}
coef(est_RIF)
coef(est_RIFINH)
coef(est_RPTINH)
```

### 5.3 Model predictions vizualization

The following figure shows the estimated inter-individual variability of the
exponential decay rate:

```{r}
x_max <- .45

x_val <- seq(0, x_max, le = 512)
add_lines <- function(x, col) {
  lines(x_val, dgamma2(x_val, x["lambda"], x["sigma"]), lwd = 2, col = col)
}

plot(NA, xlim = c(0, x_max), ylim = c(0, 25),
     xlab = "decay rate (/day)", ylab = "density of probability")

add_lines(coef(est_RIF), 2)
add_lines(coef(est_RIFINH), 3)
add_lines(coef(est_RPTINH), 4)
```

A function that adds model predictions to a plot:

```{r}
add_pred <- function(x, col, t_max, alpha = .2, f = I) {
  t_val <- seq(0, t_max, le = 512)
  q050 <- f(cfu(1e4, qgamma2(.500, x["lambda"], x["sigma"]), t_val))
  q025 <- f(cfu(1e4, qgamma2(.025, x["lambda"], x["sigma"]), t_val))
  q975 <- f(cfu(1e4, qgamma2(.975, x["lambda"], x["sigma"]), t_val))
  polygon(c(t_val, rev(t_val)), c(q025, rev(q975)), border = NA, col = adjustcolor(col, alpha))
  lines(t_val, q050, col = col, lwd = 2)
}
```

The following figure shows the model-predicted expected proportions of
individuals with CFU > 1 in the population:

```{r}
t_max <- 6 * 30

plot(NA, xlim = c(0, t_max), ylim = c(0, 1e4), xlab = "time (days)", ylab = "CFU")

add_pred(coef(est_RIF), 2, t_max)
add_pred(coef(est_RIFINH), 3, t_max)
add_pred(coef(est_RPTINH), 4, t_max)
```

Same figure on a log scale:

```{r}
plot(NA, xlim = c(0, t_max), ylim = log10(c(1e-4, 1e4)), xlab = "time (days)", ylab = "CFU")

add_pred(coef(est_RIF), 2, t_max, f = log10)
add_pred(coef(est_RIFINH), 3, t_max, f = log10)
add_pred(coef(est_RPTINH), 4, t_max, f = log10)

abline(h = 0)
```

A function that adds model-predicted proportions of individual with CFU > 1 in 
a population of 15 individuals:

```{r}
add_mod_pred <- function(x, col, alpha = .2) {
  t_val <- seq(0, 6 * 30, le = 512)
  p_val <- map_dbl(t_val, ~ probability(x["lambda"], x["sigma"], N0 = 1e4, t = .x, epsilon = 1e-16))
  t_val <- t_val / 30
  q025 <- qbinom(.025, 15, p_val) / 15
  q975 <- qbinom(.975, 15, p_val) / 15
  polygon(c(t_val, rev(t_val)), c(q025, rev(q975)), border = NA, col = adjustcolor(col, alpha))
  lines(t_val, qbinom(.5, 15, p_val) / 15, col = col)
}
```

Proportion of individuals with CFU > 1 as a function of time in a population of
15 individuals:

```{r}
treatments <- unique(table2$group)

plot(NA, xlim = c(0, 6), ylim = 0:1,
     xlab = "duration of treatment (months)",
     ylab = "proportion positive")

walk2(treatments, 1:4, add_data)

add_mod_pred(coef(est_RIF), 2)
add_mod_pred(coef(est_RIFINH), 3)
add_mod_pred(coef(est_RPTINH), 4)
```


## 6 Speculating on effects in humans


```{r}
add_pred2 <- function(x, cfu0 = 1e4, col, t_max, alpha = .2, f = I) {
  t_val <- seq(0, t_max, le = 512)
  q050 <- f(cfu(cfu0, qgamma2(.500, x["lambda"], x["sigma"]), t_val))
  q025 <- f(cfu(cfu0, qgamma2(.025, x["lambda"], x["sigma"]), t_val))
  q975 <- f(cfu(cfu0, qgamma2(.975, x["lambda"], x["sigma"]), t_val))
  polygon(c(t_val, rev(t_val)), c(q025, rev(q975)), border = NA, col = adjustcolor(col, alpha))
  lines(t_val, q050, col = col, lwd = 2)
}
```

```{r}
CFU0_effect <- function(cfu0) {
  plot(NA, xlim = c(0, t_max), ylim = log10(c(1e-4, 1e4)), xlab = "time (days)", ylab = "CFU")
  
  add_pred2(coef(est_RIF),    cfu0, 2, t_max, f = log10)
  add_pred2(coef(est_RIFINH), cfu0, 3, t_max, f = log10)
  add_pred2(coef(est_RPTINH), cfu0, 4, t_max, f = log10)
  
  abline(h = 0)
}
```

```{r}
CFU0_effect(1e4)
CFU0_effect(1e3)
CFU0_effect(1e2)
```

```{r}
cfu1_time <- function(x, cfu0 = 1e4, p = .5) {
  optimize(function(t) abs(log(cfu(cfu0, qgamma2(p, x["lambda"], x["sigma"]), t))), c(0, 200))$minimum
}
```

Let's try it:

```{r}
cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .5)
cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .025)
cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .975)
```

```{r}
CFU0_effect_cfu1_time <- function(x, cfu0 = 1e4) {
  cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .5)
  cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .025)
  cfu1_time(coef(est_RIF), cfu0 = 1e4, p = .975)
}
```

```{r}
f <- function(x, cfu0) {
  map_dfc(c(upper = .025, median = .5, lower = .975), cfu1_time, x = x, cfu0 = cfu0)
}
```

Let's try it:

```{r}
f(coef(est_RIF), 1e4)
```

```{r}
cfu0_val <- seq(1e2, 1e4, le = 512)
out <- map_dfr(cfu0_val, f, x = coef(est_RIF))
```

```{r}
add_speculation1 <- function(x, col, alpha = .2) {
  cfu0_val <- 10^seq(2, 4, le = 512)
  out <- map_dfr(cfu0_val, f, x = coef(x))
  polygon(c(cfu0_val, rev(cfu0_val)), c(out$upper, rev(out$lower)),
          border = NA, col = adjustcolor(col, .2))
  lines(cfu0_val, out$median, col = col, lwd = 3)
}
```

```{r}
plot(NA, xlim = 10^c(2, 4), ylim = c(0, 200), log = "x",
     xlab = "initial CFU", ylab = "time of clearance (days)")
add_speculation1(est_RIF, 2)
add_speculation1(est_RIFINH, 3)
add_speculation1(est_RPTINH, 4)
```

