---
title: "Of mice and men"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      message = FALSE)
```


Looking at data from
Zhang *et al.*
Short-course chemotherapy with TMC207 and rifapentine in a murine model of
latent tuberculosis infection.
*American Journal of Respiratory and Critical Care Medicine* 184.6 (2011): 732-737.
[10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)

## Packages

```{r}
library(readxl)
library(dplyr)
library(purrr)
library(tidyr)
```


## Utilitary functions

A function that add proportions estimate and confidence interval to a data frame
from a column of successes and a column or trials:

```{r}
add_prop_est <- function(df, x, n, ...) {
  df |> 
    mutate(test = map2({{ x }}, {{ n }}, prop.test, ...),
           prop = map_dbl(test, ~ .x[["estimate"]]),
           conf = map(test, ~ setNames(.x[["conf.int"]], c("lwr", "upr")))) |> 
    unnest_wider(conf) |> 
    select(- test)
}
```

```{r}
add_prop_est <- function(df, x, n, p = "est", l = "lwr", u = "upr", ...) {
  df |> 
    mutate(test   = map2({{ x }}, {{ n }}, prop.test, ...),
           "{p}" := map_dbl(test, ~ .x[["estimate"]]),
           conf   = map(test, ~ setNames(.x[["conf.int"]], c(l, u)))) |> 
    unnest_wider(conf) |> 
    select(- test)
}
```


## Loading the data

```{r}
table2 <- "table 2 Zhang et al 2011.xlsx" |>
  read_excel() |> 
  add_prop_est(positive, total)
```

The data look like this:

```{r}
table2
```


## Analysis

Adding proportions and confidence intervals:



